---
title: "ERA5 Data"
output: html_notebook
---

# Baysian GLMM Wind Energy Forecasting

By: *Aleksanteri Hämäläinen*


**Data:**

* Reanalysis data used is ERA5 data from [here](https://cds.climate.copernicus.eu)
* Wind energy data for Finland from [Fingrid](https://www.fingrid.fi/sahkomarkkinat/sahkomarkkinainformaatio/tuulivoiman-tuotanto/)

## Setup

```{r}
library(ncdf4)
library(ncdf4.helpers)
library(lubridate)
library(tidyverse)
library(rstanarm)
library(bayesplot)
```


### Library options:

```{r}
# rstanarm options
options(mc.cores = parallel::detectCores())
# Bayesplot
theme_set(bayesplot::theme_default())
```


### Target

Load and format target

```{r}
# Load
energy_2018 <- read_csv("data/energy/events_2018.csv")
energy_2019 <- read_csv("data/energy/events_2019.csv")

# Combine
target <- bind_rows(energy_2018, energy_2019) %>%
  # Drop unnecessary columns and simplify column names
  select(datetime="Alkuaika UTC", energy="Tuulivoimatuotanto - tuntienergiatieto") %>%
  mutate(
    # This is to fix some times being 5min early, e.g. 22:55 instead of 23:00
    datetime = force_tz(ceiling_date(datetime, unit="hour"), tz="UTC"),
    # Add log(energy) column
    log_energy = log(energy)
  ) %>% 
  distinct(datetime, .keep_all = TRUE) # Some timesteps have both 22:55 and 23:00 with slightly different energy values

target
```


Missing timesteps in energy data

```{r}
start <- ymd_hms("2017-12-31 22:00:00")
ts <- c(start)
for (i in 1:17522) {
  ts <- append(ts, tail(ts, n=1) + dminutes(60))
}

which(!(ts[1:17519] %in% target$datetime))
```


### Predictors

Load data and extract features

```{r}
data_filepath <- paste0("data/ecmwf/", "era5_2018-2019.nc")
data_output <- nc_open(data_filepath)

print(data_output)

feature_num <- 8 # Disregarding lat, lon and time and adding w10 and w100
u100 <- ncvar_get(data_output, varid="u100")
v100 <- ncvar_get(data_output, varid="v100")
u10 <- ncvar_get(data_output, varid="u10")
v10 <- ncvar_get(data_output, varid="v10")
t2m <- ncvar_get(data_output, varid="t2m")
sp <- ncvar_get(data_output, varid="sp")
lon <- ncvar_get(data_output, varid="longitude")
lat <- ncvar_get(data_output, varid="latitude")
time <- ncvar_get(data_output, varid="time")

nc_close(data_output)
```

Reanalysis data dimension check

```{r}
dim(u100)
```


Create w10 and w100

```{r}
w10 <- array(dim = c(length(lon), length(lat), length(time)))
w100 <- array(dim = c(length(lon), length(lat), length(time)))
for (t in 1:length(time)) {
  w10[,,t] <- sqrt(u10[,,t]**2 + v10[,,t]**2)
  w100[,,t] <- sqrt(u100[,,t]**2 + v100[,,t]**2)
}
```


Create region masks

```{r}
# Regions (17)
region_num <- 17
regions <- rbind(
    c(19.0, 21.2, 60.6, 59.5), # Ahvenanmaa
    c(21.2, 23.5, 61.0, 59.5), # Varsinais-Suomi
    c(23.5, 26.2, 60.7, 59.5), # Uusimaa
    c(26.2, 27.2, 61.3, 59.0), # Kymenlaakso
    c(27.2, 30.0, 61.7, 59.0), # Etelä-Karjala
    c(20.0, 22.7, 61.3, 60.0), # Satakunta
    c(23.5, 25.0, 61.2, 60.7), # Kanta-Häme
    c(25.0, 26.2, 61.6, 60.7), # Päijät-Häme
    c(22.7, 25.0, 62.4, 61.0), # Pirkanmaa
    c(26.2, 29.0, 62.4, 61.3), # Etelä-Savo
    c(24.5, 26.5, 63.5, 61.6), # Keskisuomi
    c(20.0, 24.5, 64.0, 61.0), # Etelä + Keski + Pohjanmaa
    c(28.4, 32.0, 63.9, 61.7), # Pohjois-Karjala
    c(26.5, 28.4, 64.0, 62.4), # Pohjois-Savo
    c(27.2, 30.5, 65.3, 63.9), # Kainuu
    c(23.4, 30.0, 66.0, 63.5), # Pohjois-Pohjanmaa
    c(22.7, 30.5, 69.5, 66.0)  # Lappi
)

# Region map base
region_map <- replicate(length(lat), numeric(length(lon))) - 1

# Fit
for (y in 1:length(lat)) {
  for (x in 1:length(lon)) {
    for (i in 1:length(regions[,1])) {
      r <- regions[i,]
      if (lon[x] >= r[1] && lon[x] <= r[2] && lat[y] <= r[3] && lat[y] >= r[4]) {
        region_map[x, y] <- i
        break
      }
    }
  }
}
```

```{r}
region_masks <- array(dim = c(length(lon), length(lat), length(regions)))
for (r in 1:length(regions)) {
  region_masks[,,r] <- region_map == r
}
```


Calculate region averages

```{r}
region_means <- array(dim = c(length(time) * region_num, feature_num + 2)) # +2 for datetime and region column
colnames(region_means) <- c("datetime", "u10", "v10", "w10", "u100", "v100", "w100", "t2m", "sp", "region")
for (t in 1:length(time)) {
  for (r in 1:region_num) {
    r_mask <- region_masks[,,r]
    region_means[(t-1) * region_num + r,] <- c(
      with_tz(ts[t + 2], tz="UTC"), # Datetime
      mean(u10[,,t][r_mask]),
      mean(v10[,,t][r_mask]),
      mean(w10[,,t][r_mask]),
      mean(u100[,,t][r_mask]),
      mean(v100[,,t][r_mask]),
      mean(w100[,,t][r_mask]),
      mean(t2m[,,t][r_mask]),
      mean(sp[,,t][r_mask]),
      r # Region column
    )
  }
}
head(region_means, n=20)
```
```{r}
# Helper to to map months to seasons
seasons <- rbind(
              c(12, 1, 2), # 1, winter
              c(3, 4, 5),  # 2, spring
              c(6, 7, 8),  # 3, summer
              c(9, 10, 11) # 4, fall
            )

season_map <- function(m) {
  for (s in 1:4) {
    if (m %in% seasons[s,]) {
      return(s)
    }
  }  
  return(-1)
}

predictors <- as_tibble(region_means) %>%
                    # Change datetime format
                    mutate(datetime = as_datetime(datetime)) %>% 
                    # Add season
                    mutate(season = map_int(month(datetime), season_map), .before = "u10")
predictors
```


### Combine

```{r}
full_df <- inner_join(target, predictors, by = "datetime")

full_df
```

### Full aggregate DF

```{r}
aggregate_df <- full_df %>% group_by(datetime) %>% 
                  mutate(
                    u10 = mean(u10),
                    v10 = mean(v10),
                    w10 = mean(w10),
                    u100 = mean(u100),
                    v100 = mean(v100),
                    w100 = mean(w100),
                    t2m = mean(t2m),
                    sp = mean(sp),
                    region = NULL
                  ) %>% 
                  distinct() %>% 
                  ungroup()
aggregate_df
```

### [Helper] Forecast Timestep Shift

```{r}
# df:      dataframe to shift
# ts:      timestep to shift in hours
# regions: number of regions
ts_shift <- function (df, ts, regions) {
  df_shift <- df %>% mutate(
                      energy = lead(df$energy, n = ts * regions),
                      log_energy = lead(df$log_energy, n = ts * regions)
                    ) %>% 
                    drop_na()
  return(df_shift)
}
```

```{r}
ag_df_s2 <- ts_shift(aggregate_df, ts=2, regions=1)
ag_df_s2
```

```{r}
full_df_ts2 <- ts_shift(full_df, ts=2, regions=17)
full_df_ts6 <- ts_shift(full_df, ts=6, regions=17)
full_df_ts24 <- ts_shift(full_df, ts=24, regions=17)

save(full_df, full_df_ts2, full_df_ts6, full_df_ts24, file = "full_df.rda")
```

### [Helper] Training data sampler

```{r}
# df:     dataframe to sample from
# years:  vector of years to include
# months: vector of months to include
# dhour:  which hours to pick (hour %% dhour == 0)
training_data_sampler <- function(df, years, months, dhour) {
  sample <- df %>% filter(
                    year(datetime) %in% years &
                    month(datetime) %in% months &
                    hour(datetime) %% dhour == 0
                  )
  return(sample)
}
```











## Models

Models fitted with `rstanarm`.

List of models:

1. Full aggregate models with forecast ts = 0:
1.1. Only w10
1.2. Only w100



### Regional w10 and w100 Predictors and Log Energy Target

**Info**

Training data: 2018, 01-02, dhour=9
Intercept prior: `student's_t(2, 4, 1)`



**Fit**

```{r}
subset_df <- training_data_sampler(full_df, c(2018), c(1, 2), 9)

post_w10_w100_log <- stan_glmer(log_energy ~ w10 + w100 + (w10 + w100 | region),
                      data = subset_df,
                      family = gaussian(link = "identity"),
                      prior_intercept = student_t(df = 2, loc = 4, scale = 1),
                      adapt_delta = 0.99
                    )

# Save the fit so it doesn't have to be run again every time RStudio dies
save(post_w10_w100_log, file = "post_w10_w100_log.rda")
# Short summary
post_w10_w100_log
```

```{r}
load("post_w10_w100_log.rda")
```




**Divergence plot**

```{r}
pairs(post_w10_w100_log, pars = c("(Intercept)", "w10", "w100"))
```



**Prior Summary**

```{r}
prior_summary(object = post_w10_w100_log)
```



**Population intercept plot**

```{r}
draws_w10_w100_log <- as.data.frame(post_w10_w100_log)

ggplot(data = draws_w10_w100_log) +
  geom_histogram(aes(x = `(Intercept)`), fill = "steelblue", color = "black")
```

**Coefficient plots**

```{r}
posterior <- as.matrix(post_w10_w100_log)
mcmc_areas(posterior,
           pars = c("w10", "w100"),
           prob = 0.8)
```



**Posterior predictions**

```{r}
nd <- tail(full_df, n = 17) %>% select(w10, w100, region)
nd_targets <- tail(full_df, n = 1) %>% select(energy)

post_predict <- posterior_predict(post_w10_w100_log, nd, draws = 500)
```



**Prediction error distribution**

```{r}
post_predict <- as_tibble(post_predict) %>%
  mutate(
    # Real target
    R = nd_targets$energy,
    # Prediction means (exp to get MWh)
    E = rowMeans(exp(.[,1:17])),
    .before = `1`) %>% 
  mutate(error = R - E, .before = `1`)
post_predict
```

```{r}
mcmc_areas(post_predict,
           pars = c("error"),
           prob = 0.5) +
  scale_x_continuous(breaks = c(-1000, -750, -500, -250, -100, 0, 100, 250, 500, 750))
  xlab("Error [MWh]")
```



**Full coefficient summary**

```{r}
summary(post_w10_w100_log)
```



### Regional w10 predictor and log energy target

**Info**

Training data: 2018, 01-03, dhour=12



**Fit**

```{r}
subset_df <- training_data_sampler(full_df, c(2018), c(1, 2), 9)

post_w10_log <- stan_glmer(log_energy ~ w10 + (w10 | region),
                data = subset_df,
                family = gaussian(link = "identity"),
                prior_intercept = student_t(2, 4, 1),
                adapt_delta = 0.99
                )

save(post_w10_log, file = "post_w10_log.rda")

# Short summary
post_w10_log
```

```{r}
load("post_w10_log.rda")
```

```{r}
pairs(post_w10_log, pars = c("(Intercept)", "log-posterior"))
```



**Prior Summary**

```{r}
prior_summary(object = post_w10_log)
```



**Population intercept plot**

```{r}
draws_w10_log <- as.data.frame(post_w10_log)

ggplot(data = draws_w10_log) +
  geom_histogram(aes(x = `(Intercept)`, y = after_stat(density)), fill = "steelblue", color = "black") +
  xlim(3.8, 4.8)
```

**Coefficient plots**

```{r}
posterior <- as.matrix(post_w10_log)
mcmc_areas(posterior,
           pars = c("w10"),
           prob = 0.8)
```



**Posterior predictions**

```{r}
nd <- tail(full_df, n = 17) %>% select(w10, region)
nd_targets <- tail(full_df, n = 1) %>% select(energy)

pp_w10_log <- posterior_predict(post_w10_log, nd, draws = 500)
```



**Prediction error distribution**

```{r}
pp_w10_log <- as_tibble(pp_w10_log) %>%
  mutate(
    # Real target
    real_target = nd_targets$energy,
    # Prediction means (exp to get MWh)
    predicted_target = rowMeans(exp(.[,1:17])),
    .before = `1`) %>% 
  mutate(
    error = real_target - predicted_target,
    .after = predicted_target)

pp_w10_log
```

```{r}
mcmc_areas(pp_w10_log,
           pars = c("error"),
           prob = 0.8) +
  xlab("Error [MWh]")
```



**Full coefficient summary**

```{r}
summary(post_w10_log)
```








### Regional w10 and w100 Predictors, Log Energy Target and 2h timeshift

**Info**

Training data: 2018, 01-02, dhour=9
Intercept prior: `student's_t(2, 4, 1)`
Time shift: 2h


**Fit**

```{r}

ts2_df <- ts_shift(
  training_data_sampler(full_df, c(2018), c(1, 2), 9),
  ts=2,
  regions=17)

post_w10_w100_log_ts2 <- stan_glmer(log_energy ~ w10 + w100 + (w10 + w100 | region),
                      data = ts2_df,
                      family = gaussian(link = "identity"),
                      prior_intercept = student_t(df = 2, loc = 4, scale = 1),
                      adapt_delta = 0.999
                    )

# Save the fit so it doesn't have to be run again every time RStudio dies
save(post_w10_w100_log_ts2, file = "post_w10_w100_log_ts2.rda")
# Short summary
post_w10_w100_log_ts2
```



**Prior Summary**

```{r}
prior_summary(object = post_w10_w100_log_ts2)
```



**Population intercept plot**

```{r}
draws_w10_w100_log_ts2 <- as.data.frame(post_w10_w100_log_ts2)

ggplot(data = draws_w10_w100_log_ts2) +
  geom_histogram(aes(x = `(Intercept)`), fill = "steelblue", color = "black")
```

**Coefficient plots**

```{r}
posterior <- as.matrix(post_w10_w100_log_ts2)
mcmc_areas(posterior,
           pars = c("w10", "w100"),
           prob = 0.8)
```



**Posterior predictions**

```{r}
nd <- tail(ts2_df, n = 17) %>% select(w10, w100, region)
nd_targets <- tail(ts2_df, n = 1) %>% select(energy)

pp_w10_w100_log_ts2 <- posterior_predict(post_w10_w100_log_ts2, nd)
```



**Prediction error distribution**

```{r}
pp_w10_w100_log_ts2 <- as_tibble(pp_w10_w100_log_ts2) %>%
  mutate(
    # Real target
    real_target = nd_targets$energy,
    # Prediction means (exp to get MWh)
    predicted_target = rowMeans(exp(.[,1:17])),
    .before = `1`) %>% 
  mutate(
    error = real_target - predicted_target,
    .after = predicted_target)

pp_w10_w100_log_ts2
```

```{r}
mcmc_areas(pp_w10_w100_log_ts2,
           pars = c("error"),
           prob = 0.8) +
  xlab("Error [MWh]")
```



**Full coefficient summary**

```{r}
summary(post_w10_w100_log_ts2)
```



### Regional w10, w100 and t2m Predictors, Log Energy Target and 2h timeshift

**Info**

Training data: 2018, 01-02 + 11, dhour=6
Intercept prior: `student's_t(2, 4, 1)`
Time shift: 2h


**Fit**

```{r}
ts2_df <- ts_shift(
  training_data_sampler(full_df, c(2018), c(11, 1, 2), 6),
  ts=2,
  regions=17)

post_w10_w100_t2m_log_ts2 <- stan_glmer(log_energy ~ w10 + w100 + t2m + (w10 + w100 + t2m | region),
                      data = ts2_df,
                      family = gaussian(link = "identity"),
                      prior_intercept = student_t(df = 2, loc = 4, scale = 1),
                      adapt_delta = 0.999
                    )

# Save the fit so it doesn't have to be run again every time RStudio dies
save(post_w10_w100_t2m_log_ts2, file = "post_w10_w100_t2m_log_ts2.rda")
# Short summary
post_w10_w100_t2m_log_ts2
```


```{r}
mcmc_areas(post_w10_w100_t2m_log_ts2,
           pars = c("w10", "w100"),
           prob = 0.8)
```

### Regional w10, w100 and sp Predictors, Log Energy Target and 2h timeshift [summer]

**Info**

Training data: 2018, 06-08, dhour=6
Intercept prior: `student's_t(2, 4, 1)`
Time shift: 2h


**Fit**

```{r}
ts2_df_summer <- ts_shift(
  training_data_sampler(full_df, c(2018), c(6, 7, 8), 6),
  ts=2,
  regions=17)

post_w10_w100_sp_log_ts2_summer <- stan_glmer(log_energy ~ w10 + w100 + sp + (w10 + w100 + sp | region),
                      data = ts2_df_summer,
                      family = gaussian(link = "identity"),
                      prior_intercept = student_t(df = 2, loc = 4, scale = 1),
                      adapt_delta = 0.999
                    )

# Save the fit so it doesn't have to be run again every time RStudio dies
save(post_w10_w100_sp_log_ts2_summer, file = "post_w10_w100_sp_log_ts2_summer.rda")
# Short summary
post_w10_w100_sp_log_ts2_summer
```

```{r}
mcmc_areas(post_w10_w100_sp_log_ts2_summer,
           pars = c("w10", "w100"),
           prob = 0.8)
```

```{r}
mcmc_areas(post_w10_w100_sp_log_ts2_summer,
           pars = c("sp"),
           prob = 0.8)
```



## w10, w100, t2m, sp, log energy, and timeshift 2

```{r}
ts2_df <- ts_shift(
  training_data_sampler(full_df, c(2018), c(11, 1, 2), 6),
  ts=2,
  regions=17)

post_w10_w100_t2m_sp_log_ts2 <- stan_glmer(log_energy ~ w10 + w100 + t2m + sp + (w10 + w100 + t2m + sp | region),
                      data = ts2_df,
                      family = gaussian(link = "identity"),
                      prior_intercept = student_t(df = 2, loc = 4, scale = 1),
                      adapt_delta = 0.999
                    )

# Save the fit so it doesn't have to be run again every time RStudio dies
save(post_w10_w100_t2m_sp_log_ts2, file = "post_w10_w100_t2m_sp_log_ts2.rda")
# Short summary
post_w10_w100_t2m_sp_log_ts2
```

```{r}
mcmc_areas(post_w10_w100_t2m_sp_log_ts2,
           pars = c("sp"),
           prob = 0.8)
```

```{r}
mcmc_areas(post_w10_w100_t2m_sp_log_ts2,
           pars = c("t2m"),
           prob = 0.8)
```

```{r}
mcmc_areas(post_w10_w100_t2m_sp_log_ts2,
           pars = c("w10", "w100"),
           prob = 0.8)
```


































### Full Aggregate with w10 Predictor and Energy Target

```{r}
post_ag_w10 <- stan_glm(energy ~ w10,
                data = aggregate_df,
                family = gaussian(link = "identity"))
```

```{r}
draws_ag_w10 <- as.data.frame(post_ag_w10)
colnames(draws_ag_w10) <- c("a", "b0", "sigma")

post_ag_w10
```

```{r}
base <- ggplot(aggregate_df, aes(x = w10, y = energy)) +
         geom_point(size=1)

base
  # geom_abline(data = draws_ag_w10, aes(intercept = a, slope = b0),
              # color = "skyblue", size = 0.2, alpha = 0.25) +
  # geom_abline(intercept = coef(post_ag_w10)[1], slope = coef(post_ag_w10)[2],
              # color = "skyblue4", size = 1)
```


### Full Aggregate with w10 Predictor and Log Energy Target

```{r}
post_ag_w10_log <- stan_glm(log_energy ~ w10,
                    data = aggregate_df,
                    family = gaussian(link = "identity"))
```

```{r}
draws_ag_w10_log <- as.data.frame(post_ag_w10_log)
colnames(draws_ag_w10_log) <- c("a", "b0", "sigma")

post_ag_w10_log
```

```{r}
base_log <- ggplot(aggregate_df, aes(x = w10, y = log_energy)) +
              geom_point(size=1) +
              scale_y_continuous(trans = "log")
base_log +
#  scale_y_continuous(trans = "log") +
  geom_abline(data = draws_ag_w10_log, aes(intercept = a, slope = b0),
              color = "skyblue", size = 0.2, alpha = 0.25) +
  geom_abline(intercept = coef(post_ag_w10_log)[1], slope = coef(post_ag_w10_log)[2],
              color = "skyblue4", size = 1)
```

```{r}
base_log <- ggplot(aggregate_df, aes(x = w10, y = log_energy)) +
              geom_point(size=1)
base_log +
#  scale_y_continuous(trans = "log") +
  geom_abline(data = draws_ag_w10_log, aes(intercept = a, slope = b0),
              color = "skyblue", size = 0.2, alpha = 0.25) +
  geom_abline(intercept = coef(post_ag_w10_log)[1], slope = coef(post_ag_w10_log)[2],
              color = "skyblue4", size = 1)
```













